Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-08-22 doc/invt.tex
invt.tex: small edits

\documentclass[preprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{kim2006, *kim2007, kim2010, junghans2014,
  langfeld2012, pellegrini2014,
  maragliano2006, *abrams2008,
  zheng2010}.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Often, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.\cite{junghans2014}
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations\cite{wang2001,
  wang2001pre, kim2006, *kim2007},
the latter is more popular for MD simulations
as it provides a smoother bias potential
that can be readily differentiated for force calculation.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
In both cases,
regular updates to the bias potential
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one needs to gradually decrease
the updating magnitude of the bias.

The schedule of reducing the updating magnitude
can critically affect the precision of the final bias potential,
and hence that of the PMF\cite{laio2005, bussi2006, poulain2006,
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007, min2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
Overly slow and rapid reduction of the updating magnitude
can hurt the precision and accuracy of the bias potential.
%
For the single-bin case, it is known that
the optimal updating schedule is given by the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of updating steps.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we briefly review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we derive the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
We then compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.
%
Finally,
we use the bandpass updating schemes
to build a class of adaptive FDS algorithms
with quadratic bias potentials.

@@

We will first review the spectral decomposition
of the updating matrix, $\mathbf w$,
which helps establish the convergence criterion
and the decomposition of error into
independent eigenmodes.
%
The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
If the rate of growth is positive,
we can, by a proper overall scaling of $\mathbf w$,
write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will %take advantage of the analogy and
borrow techniques used
in studying transition matrices\cite{vankampen},
but limit ourselves to properties that are unaffected
by the admission of negative elements into $\mathbf w$.

@@

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
we have chosen the $\mathcal N+1$ grid points
($\mathcal N=20000$ here)
over $[0, q(T)]$ to be given by
$q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
where $i = 0, \dots, \mathcal N$.\footnote{For
  a mass distribution, $m(\bar q)$, that is
  inversely proportional to $\bar q$,
  this grid would make the integral,
  or the difference, $t_{[i+1]} - t_{[i]}$,
  roughly a constant.}
%
The curve $q(t)$ is then integrated
using the trapezoidal rule\cite{press3rd},
resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude
from (cf. Appendix \ref{sec:ave_us}):
\begin{equation}
  \alpha_i = \frac{ 1 - e^{-\lambda_1 \, \hat \alpha_i} } { \lambda_1 }
  ,
  \label{eq:alphai}
\end{equation}
where
$\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv 0$,
$q_{[\mathcal N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[\mathcal N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linearly interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

We will first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of updating matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s\footnote{This
inequality follows from the nonnegativity of
the quadratic polynomial,
$$
\int_0^T
  dt \sum_{k = 1}^{n-1} \Gamma_k \,
    \left( {\dot \vartheta}_k \, \chi - c_k \right)^2
  \equiv
  A_2 \, \chi^2 + A_1 \, \chi + A_0
  ,
$$
which implies a nonpositive discriminant,
$A_1^2 - 4 \, A_0 \, A_2$.}
%
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, {\dot \vartheta}_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 1}^{n-1}
      \Gamma_k \, c_k \, {\dot \vartheta}_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 1}^{n-1} \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\notag
%\label{eq:CSineq}
\end{align}
%
Note that the last expression %of Eq. \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err_A(T)$
given in Eq. \eqref{eq:error_asym}.
%
The equality is achieved
if $\dot \vartheta_k\left( q(t) \right) = c_k$
for all $k > 0$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
a set of identical eigenvalues,
$\lambda_1 = \cdots = \lambda_{n-1}$
and identical $c_k$'s.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that discard certain noise modes
of short wavelengths.
%
If the eigenmodes are sorted in descending order
of the wavelength,
and the PMF is sufficiently smooth
to be representable by the first $K$
eigenmodes,
we may assign the first $K$ eigenmodes with eigenvalue $1$
and the rest of eigenmodes with eigenvalue $0$:
%
\begin{equation}
  \lambda_1 = \cdots = \lambda_{K-1} = 1
  ,
  \mathrm{\; and \;}
  \lambda_K = \cdots = \lambda_{n-1} = 0
  .
  \label{eq:lambda_bandpass}
\end{equation}
%
The updating matrix can then be reconstructed
from the eigenvalues by Eq. \eqref{eq:w_from_phi}.
%(cf. Appendix \ref{sec:homo_bandpass} for the case of homogeneous updating schemes).
%
%Then, the updating matrix is, according to
%Eq. \eqref{eq:w_from_phi},
%%
%$$w_{ij} = \frac{1}{\rho_i} \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj}.$$
%
In effect, each updating step of the bandpass updating scheme
is equivalent to a single-bin updating step,
Eq. \eqref{eq:wl_update},
followed by a filtering step
that completely removes the components of the noise modes of $k \ge K$
from the bias potential.
%
%The asymptotic error is then
%$\Err_A(T) = \sum_{ k = 1 }^{ K - 1 } \Gamma_k T/(T + t_0)^2$,
%which is less than that from Eq. \eqref{eq:EA_sbin1}.

Applying this construction to the basis of cosine and sine functions,
we get a homogeneous bandpass updating scheme
that resembles the Gaussian updating scheme,
but with better convergence in the asymptotic limit.
%
In Sec. \ref{sec:lj},
we will further compare the performance of the two updating schemes
in an example.
%
%While both updating scheme are able to deliver
%a smooth and differentiable potential of mean force,
%the bandpass updating scheme is superior in asymptotic convergence,
%because of the set of clear-cut spectrum of eigenvalues.
%Eq. \eqref{eq:lambda_bandpass},
%instead of a Gaussian one, Eq. \eqref{eq:lambda_Gaussian},
%
%\footnote{As an estimate of error,
%  we note that truncating a linear or quadratic function
%  at mode $K$ for a non-periodic variable
%  results in an error roughly proportional to
%  $|\max_i u_i - \min_i u_i|^2/K^3$.}

@@

In addition to the above application to
homogeneous updating schemes,
we will consider here a family of bandpass updating schemes
constructed from a sequence of orthogonal polynomials.
%
These updating schemes are more intuitive and easier to implement
as the bias potential in this case is
a low-order polynomial of the collective variable.

@@

It is convenient to
replace the discrete bin index, $i$,
by the continuous variable, $z$;
%
and we will replace $\rho_i$ by $\rho(z) \, \delta z$,
  $\phi_{ki}$ by $\phi_k(z) \, \delta z$,
  and
  $w_{ij}$ by $w(z, z') \, \delta z$, etc.
%
We define the eigenvectors by
$\phi_k(z) = R_k(z) \, \rho(z)$,
where $R_k(z)$ is the $k$th member of an sequence of orthogonal polynomials
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z = \delta_{k, k'}
  .
\end{align*}
%
We can readily verify that the eigenvectors satisfy
Eqs. \eqref{eq:eig_orthonormal_cols}, \eqref{eq:eig_orthonormal_rows},
and \eqref{eq:ortho0}.
%
\note{For $k = 0$, $R_0(z)$ must be a constant,
  and by the orthonormal condition, we have $R_0(z) = 1$.
  Thus, Eq. \eqref{eq:ortho0} is satisfied.}%
%
The eigenvectors allow us to construct the updating matrix
from Eq. \eqref{eq:w_from_phi},
with the eigenvalues given by \eqref{eq:lambda_bandpass}:
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to a uniform shift
of the bias potential.
%
By Eq. \eqref{eq:mbin_update},
the bias potential is updated as
%
$$
u(z, t+1) = u(z, t)
+ \sum_{k=1}^{K-1} R_k(z) R_k\bigl( z(t) \bigr) \, \alpha(t).
$$
%
We can write the bias potential
as a superposition of the orthogonal polynomials
%
\begin{equation}
  u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z),
  \label{eq:uz_decomp}
\end{equation}
%
with the coefficients updated as
\begin{equation}
  c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr)
  .
  \label{eq:ckupdate}
\end{equation}

For $k > 0$, $R_k(z)$ is a polynomial of degree $k$ with zero expectation.
%
Thus, $R_k\bigl(z(t)\bigr)$ roughly
represents the deviation of the $k$th statistical moment
of the instantaneous distribution
from its expectation.
Since high-order moments are generally less reliable,
this method is best used with a small $K$ for a narrow distribution.
%
However, by constructing a mixture\cite{shirts2017}
or expanded ensemble\cite{swendsen1986,
  *geyer1991, *hukushima1996, *hansmann1997, *sugita1999,
  *earl2005, *zuckerman2011, *rauscher2009,
  neuhaus2006, *neuhaus2007, kim2010,
  marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st},
we can simultaneously sample multiple target distributions
to cover a wider range, as exemplified below.

@@

For example,
to sample a flat distribution, $\rho(z) = 1/\Delta z$,
over $[z_c - \Delta z/2, z_c + \Delta z/2]$,
we can use the Legendre polynomials\cite{arfken}.
%
If the interval is sufficiently small,
we can limit the basis to a single linear mode,
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
with $K = 2$.
%
The bias potential is a linear function of $z$,
and this updating scheme is essentially the LLR algorithm\cite{langfeld2012}
for the single-interval case,
and the inverse-time schedule
was recommended for optimal convergence of $c_1$\cite{pellegrini2014}.

@@

In analogy to the LLR algorithm,
we will consider the case of sampling a Gaussian distribution
$\rho(z) \propto \exp[-(z-z_c)^2/(2\sigma_z^2)]$,
with a basis constructed from
the Hermite polynomials\cite{arfken}.
%
As discussed above,
the method is effective only to
a local region of a multimodal distribution,
and thus we will assume a small width, $\sigma_z$.
%
Because the Gaussian distribution has no hard boundaries
as in the previous case,
we need to use
in addition to the linear mode,
$R_1(z) = (z - z_c)/\sigma_z$,
a quadratic one,
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$,
to restrict the sampling region.
%
The bias potential is
quadratic\cite{maragliano2006, *abrams2008, zhu2012, neuhaus2006, *neuhaus2007},
which is common in umbrella sampling\cite{torrie1974, *torrie1977}
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006}.

@@

After simulation,
the weighted histogram analysis method
(WHAM)\cite{ferrenberg1988, *ferrenberg1989,
*kumar1992, *roux1995, *bartels1997, *souaille2001,
*kobrak2003, *gallicchio2005, *chodera2007,
*fenwick2008, *kim2011, *shirts2008,
zhu2012, newman, frenkel},
can be used to recover the unbiased distribution,
%
\begin{equation}
  p(z) \, \delta z
  =
  \frac{
    \sum_i n^{(i)}(z)
  }
  {
    \sum_i N^{(i)} \, e^{-u^{(j)}(z)} / Z^{(i)}
  }
  ,
  \notag
\end{equation}
%
where $n^{(i)}(z)$
the number of visits to the bin $(z-\delta/2, z+\delta/2)$
in umbrella $i$,
and the $N^{(i)}$ is the total number of visits
to the umbrella $i$.
%
We can use the final converged value of $Z^{(i)}$
without going through WHAM iteration.

>>>>

\documentclass[preprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
%\usepackage{tikz}
%\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, wang2001pre}
and metadynamics\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, micheletti2004}
among others\cite{yan2004, kim2006, *kim2007, kim2010, junghans2014,
  langfeld2012, pellegrini2014,
  maragliano2006, *abrams2008,
  zheng2010}.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Often, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.\cite{junghans2014}
%
We will refer to the two updating schemes
as the single-bin
and Gaussian updating schemes, respectively.
%
While the former is popular for MC simulations\cite{wang2001,
  wang2001pre, kim2006, *kim2007},
the latter is more popular for MD simulations
as it provides a smoother bias potential
that can be readily differentiated for force calculation.\cite{huber1994,
  *laio2002, *laio2008, *barducci2011, *sutto2012, junghans2014}
%
In both cases,
regular updates to the bias potential
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one needs to gradually decrease
the updating magnitude of the bias.

The schedule of reducing the updating magnitude
can critically affect the precision of the final bias potential,
and hence that of the PMF\cite{laio2005, bussi2006, poulain2006,
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007, min2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014,
barducci2008, dickson2011, dama2014}.
%
While an overly slow reduction of the updating magnitude
increases the variance of the bias potential,
an too rapid reduction often
converges the bias potential to wrong value.
%
For the single-bin case, it is known that
the optimal updating schedule is given by the
inverse time\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008}
or
the inverse number of updating steps.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we briefly review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we derive the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
We then compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.
%
Finally,
we use the bandpass updating schemes
to build a class of adaptive FDS algorithms
with polynomial bias potentials
in Sec. \ref{sec:bandpass_poly}.

@@

We will first review the spectral decomposition
of the updating matrix, $\mathbf w$,
which helps establish the convergence criterion
and the decomposition of error into
independent eigenmodes.
%
The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
If the rate of growth is positive,
we can, by a proper overall scaling of $\mathbf w$,
write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
for our cases here.
%
Below we will %take advantage of the analogy and
borrow techniques used
in studying transition matrices\cite{vankampen},
but limit ourselves to properties that are unaffected
by the admission of negative elements into $\mathbf w$.

@@

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
we have chosen the $\mathcal N+1$ grid points
($\mathcal N=20000$ here)
over $[0, q(T)]$ to be given by
$q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
where $i = 0, \dots, \mathcal N$.\footnote{For
  a mass distribution, $m(\bar q)$, that is
  inversely proportional to $\bar q$,
  this grid would make the integral,
  or the difference, $t_{[i+1]} - t_{[i]}$,
  roughly a constant.}
%
The curve $q(t)$ is then integrated
using the trapezoidal rule\cite{press3rd},
resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude
as (cf. Appendix \ref{sec:ave_us}):
\begin{equation}
  \alpha_{[i]}
  =
  \frac{1}{\lambda}
  \left[
    1
    -
    \exp\left(
    -\lambda_1
    \frac{ q_{[i+1]} - q_{[i-1]} }
         { t_{[i+1]} - t_{[i-1]} }
    \right)
  \right]
  ,
  \label{eq:alphai}
\end{equation}
with
$q_{[ -1]} \equiv 0$,
$q_{[\mathcal N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[\mathcal N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linearly interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

We will first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of updating matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s\footnote{This
inequality follows from the nonnegativity of
the quadratic polynomial,
$$
\int_0^T
  dt \sum_{k = 1}^{n-1} \Gamma_k \,
    \left( {\dot \vartheta}_k \, \chi - c_k \right)^2
  \equiv
  A_2 \, \chi^2 + A_1 \, \chi + A_0
  ,
$$
which implies a nonpositive discriminant,
$A_1^2 - 4 \, A_0 \, A_2$.}
%
%
\begin{align}
  &
  \left(
    \int_0^T dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, {\dot \vartheta}_k^2\bigl( q(t) \bigr)
  \right)
  %
  \left(
    \int_0^T dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, c_k^2
  \right)
  %
  \notag
  \\
  &
  \qquad \qquad \qquad
  \ge
  \left(
    \int_0^T dt
      \sum_{k = 1}^{n-1}
        \Gamma_k \, c_k \, {\dot \vartheta}_k \, \bigl( q(t) \bigr)
  \right)^2
  \notag
  \\
  &
  \qquad \qquad \qquad
  =
  \left(
    \sum_{k = 1}^{n-1} \Gamma_k \, c_k
      \left[
        1 - e^{ -\lambda_k \, q(T) }
      \right]
  \right)^2.
  \notag
  %\label{eq:CSineq}
\end{align}
%
Note that the last expression %of Eq. \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err_A(T)$
given in Eq. \eqref{eq:error_asym}.
%
The equality is achieved
if $\dot \vartheta_k\left( q(t) \right) = c_k$
for all $k > 0$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
a set of identical eigenvalues,
$\lambda_1 = \cdots = \lambda_{n-1}$
and identical $c_k$'s.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that discard certain noise modes
of short wavelength.
%
If the eigenmodes are sorted in descending order
of the wavelength,
and the PMF is sufficiently smooth
to be representable by the first $K$
eigenmodes,
we may assign the first $K$ eigenmodes eigenvalue $1$
and the rest of eigenmodes eigenvalue $0$:
%
\begin{equation}
  \lambda_1 = \cdots = \lambda_{K-1} = 1
  ,
  \mathrm{\; and \;}
  \lambda_K = \cdots = \lambda_{n-1} = 0
  .
  \label{eq:lambda_bandpass}
\end{equation}
%
The updating matrix can then be reconstructed
from the eigenvalues by Eq. \eqref{eq:w_from_phi}.
%(cf. Appendix \ref{sec:homo_bandpass} for the case of homogeneous updating schemes).
%
%Then, the updating matrix is, according to
%Eq. \eqref{eq:w_from_phi},
%%
%$$w_{ij} = \frac{1}{\rho_i} \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj}.$$
%
In effect, each updating step of the bandpass updating scheme
is equivalent to a single-bin updating step,
Eq. \eqref{eq:wl_update},
followed by a filtering step
that completely removes the components of the noise modes of $k \ge K$
from the bias potential.
%
%The asymptotic error is then
%$\Err_A(T) = \sum_{ k = 1 }^{ K - 1 } \Gamma_k T/(T + t_0)^2$,
%which is less than that from Eq. \eqref{eq:EA_sbin1}.

We will consider two special cases of the bandpass updating schemes.
%
First,
by applying this construction to the basis of cosine and sine functions,
we get a homogeneous bandpass updating scheme
that resembles the Gaussian updating scheme,
but with better convergence in the asymptotic limit
(cf. Appendix \ref{sec:homo_bandpass}).
%
In Sec. \ref{sec:lj},
we will further compare the performance of the two updating schemes
in an example.
%
%While both updating scheme are able to deliver
%a smooth and differentiable potential of mean force,
%the bandpass updating scheme is superior in asymptotic convergence,
%because of the set of clear-cut spectrum of eigenvalues.
%Eq. \eqref{eq:lambda_bandpass},
%instead of a Gaussian one, Eq. \eqref{eq:lambda_Gaussian},
%
%\footnote{As an estimate of error,
%  we note that truncating a linear or quadratic function
%  at mode $K$ for a non-periodic variable
%  results in an error roughly proportional to
%  $|\max_i u_i - \min_i u_i|^2/K^3$.}
%
The second case is based on a family of orthogonal polynomials,
which we will discuss below.

@@

Here we will consider here a family of bandpass updating schemes
constructed from a sequence of orthogonal polynomials.
%
These updating schemes are more intuitive and easier to implement
as the bias potential in this case is
a low-order polynomial of the collective variable.

@@

It is convenient to
replace the discrete bin index, $i$,
by the continuous variable, $z$;
%
and we will replace $\rho_i$ by $\rho(z) \, \delta z$,
  $\phi_{ki}$ by $\phi_k(z) \, \delta z$,
  and
  $w_{ij}$ by $w(z, z') \, \delta z$, etc.
%
We define the eigenvectors by
$\phi_k(z) = R_k(z) \, \rho(z)$,
where $R_k(z)$ is the $k$th member of an sequence of orthogonal polynomials
that satisfy
\begin{align*}
  \int R_k(z) \, R_{k'}(z) \, \rho(z) \, d z
  = \overline{ R_k(z) \, R_{k'}(z) }
  = \delta_{k, k'}
  ,
\end{align*}
%
where the overline denotes the average for the distribution, $\rho(z)$.
%
The first few members are
\begin{align*}
  R_0(z) = 1,
  R_1(z) = \frac{z - \overline{z}}{\sigma_z},
  R_2(z) = \frac{ (z - \overline{z})^2 - \sigma_z^2 }
  {\sqrt{\overline{\left[
    (z - \overline z)^2 - \sigma_z^2
  \right]^2 }}}
  ,
\end{align*}
%
where $\sigma_z \equiv \sqrt{\overline{\left(z - \overline z\right)^2}}$.
%
We can readily verify that the eigenvectors satisfy
Eqs. \eqref{eq:eig_orthonormal_cols}, \eqref{eq:eig_orthonormal_rows},
and \eqref{eq:ortho0}.
%
\note{For $k = 0$, $R_0(z)$ must be a constant,
  and by the orthonormal condition, we have $R_0(z) = 1$.
  Thus, Eq. \eqref{eq:ortho0} is satisfied.}%
%
The eigenvectors allow us to construct the updating matrix
from Eq. \eqref{eq:w_from_phi},
with the eigenvalues given by \eqref{eq:lambda_bandpass}:
\begin{align*}
  w(z, z')
  =
  \sum_{k=1}^{K-1} \frac{ \phi_k(z) \, \phi_k(z') } { \rho(z) }
  = \sum_{k=1}^{K-1} R_k(z) \, R_k(z') \, \rho(z'),
\end{align*}
%
where we have dropped the $k=0$ mode
that corresponds to a uniform shift
of the bias potential.
%
By Eq. \eqref{eq:mbin_update},
the bias potential is updated as
%
$$
u(z, t+1) = u(z, t)
+ \sum_{k=1}^{K-1} R_k(z) R_k\bigl( z(t) \bigr) \, \alpha(t).
$$
%
We can write the bias potential
as a superposition of the orthogonal polynomials
%
\begin{equation}
  u(z, t) = \sum_{k=1}^{K-1} c_k(t) \, R_k(z),
  \label{eq:uz_decomp}
\end{equation}
%
with the coefficients updated as
\begin{equation}
  c_k(t+1) = c_k(t) + \alpha(t) \, R_k\bigl( z(t) \bigr)
  .
  \label{eq:ckupdate}
\end{equation}

For $k > 0$, $R_k(z)$ is a polynomial of degree $k$ with zero expectation.
%
Thus, $R_k\bigl(z(t)\bigr)$ roughly
represents the deviation of the $k$th statistical moment
of the instantaneous distribution
from its expectation.
Since high-order moments are generally less reliable,
this method is better used with a small $K$ for a narrow distribution.
%
However, by constructing
an expanded ensemble or mixture\cite{swendsen1986,
  *geyer1991, *hukushima1996, *hansmann1997, *sugita1999,
  *earl2005, *zuckerman2011, *rauscher2009,
  neuhaus2006, *neuhaus2007, kim2010,
  marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st, shirts2017},
we can simultaneously sample multiple target distributions
to cover a wider range, as exemplified below.

@@

For example,
to sample a flat distribution, $\rho(z) = 1/\Delta z$,
over $\left[z_c - \frac 1 2 \Delta z, z_c + \frac 1 2 \Delta z\right]$,
we can use the Legendre polynomials\cite{arfken}.
%
If the interval is sufficiently small,
we can limit the basis to a single linear mode,
$R_1(z) = 2 \sqrt{3} \, (z - z_c)/\Delta z$,
with $K = 2$.
%
%The bias potential is a linear function of $z$,
The updating scheme is essentially the LLR algorithm\cite{langfeld2012}
for the single-interval case,
and the inverse-time schedule
was recommended for optimal convergence of $c_1$\cite{pellegrini2014}.

@@

In analogy to the LLR algorithm,
we will consider the case of sampling a Gaussian distribution
$\rho(z) \propto \exp[-(z-z_c)^2/(2\sigma_z^2)]$,
with a basis constructed from
the Hermite polynomials\cite{arfken}.
%
As discussed above,
the method is effective only for
a local region of a multimodal distribution,
and thus we will assume a small width, $\sigma_z$.
%
Because the Gaussian distribution has no hard boundaries
as in the previous case,
we need to use
in addition to the linear mode,
$R_1(z) = (z - z_c)/\sigma_z$,
a quadratic one,
$R_2(z) = \bigl[(z - z_c)^2 - \sigma_z^2\bigr] /\left(\sqrt 2 \, \sigma_z^2\right)$,
to restrict the sampling region.
%
The bias potential is
quadratic\cite{maragliano2006, *abrams2008, zhu2012, neuhaus2006, *neuhaus2007},
which is common in umbrella sampling\cite{torrie1974, *torrie1977}
and the method of Gaussian ensemble\cite{hetherington1987,
*challa1988, *costeniuc2006}.

@@

After simulation,
the weighted histogram analysis method
(WHAM)\cite{ferrenberg1988, *ferrenberg1989,
*kumar1992, *roux1995, *bartels1997, *souaille2001,
*kobrak2003, *gallicchio2005, *chodera2007,
*fenwick2008, *kim2011, *shirts2008,
zhu2012, newman, frenkel},
can be used to recover the unbiased distribution,
%
\begin{equation}
  p(z) \, \delta z
  \propto
  \frac{
    \sum_i n^{(i)}(z)
  }
  {
    \sum_i N^{(i)} \, e^{-u^{(j)}(z)} / Z^{(i)}
  }
  ,
  \notag
\end{equation}
%
where $n^{(i)}(z)$
the number of visits to the bin
$\left(z-\frac{1}{2} \delta z, z + \frac{1}{2} \delta z\right)$
in umbrella $i$,
and the $N^{(i)}$ is the total number of visits
to the umbrella $i$.
%
We can use the final converged values of $Z^{(i)}$'s
without going through the usual WHAM iteration.
